---
title: "Introduction to correlation"
output:
  beamer_presentation:
    theme: "default"
    colortheme: "default"
    fonttheme: "default"
  ioslides_presentation: default
  slidy_presentation: default
header-includes:
- \usepackage{hyperref}
- \usepackage{caption}
- \definecolor{links}{HTML}{2A1B81}
- \hypersetup{colorlinks,linkcolor=,urlcolor=links}
colorlinks: true
linkcolor: blue
urlcolor: blue
---

```{r, echo = FALSE}
library(knitr);
opts_chunk$set(echo = FALSE);
```


## Introduction to correlation

We often want to investigate the relationship between pairs of variables.

- Vegetation height and mean annual temperature
- Animal body size and metabolic rate
- Number of automobiles in a location and carbon emissions

The **correlation** between pairs of variables, such as those listed above, describes how the variation of each variable is related to the other variable.

## Visualising the correlation between two variables


```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.cap= "Two plots of hypothetical variables illustrating a positive (left) and negative correlation"}
rho1 <- 0.9;
Z1   <- rnorm(n = 16, mean = 0, sd = 1);
Z2   <- rnorm(n = 16, mean = 0, sd = 1);
X1   <- Z1 + 10;
Y1   <- rho1 * Z1 + sqrt(1 - rho1*rho1) * Z2 + 10;
par(mfrow = c(1, 2), mar = c(5, 2, 1, 0.5));
plot(x = X1, y = Y1, pch = 20, xlab = "X value", ylab = "Y value", cex = 1.5,
     xlim = c(7, 13), ylim = c(7, 13), xaxt = "n", yaxt = "n", cex.lab = 1.25,
     line = 1);
rho1 <- -0.9;
Z1   <- rnorm(n = 16, mean = 0, sd = 1);
Z2   <- rnorm(n = 16, mean = 0, sd = 1);
X1   <- Z1 + 10;
Y1   <- rho1 * Z1 + sqrt(1 - rho1*rho1) * Z2 + 10;
plot(x = X1, y = Y1, pch = 20, xlab = "X value", cex = 1.5, line = 1, 
     xlim = c(7, 13), ylim = c(7, 13), xaxt = "n", yaxt = "n", cex.lab = 1.25,
     ylab = "Y value");
```



## Visualising two variables that are not correlated

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.cap = "A plot of two hypothetical variables that are not correlated."}
X1 <- rnorm(n = 24, mean = 0, sd = 1);
Y1 <- rnorm(n = 24, mean = 0, sd = 1);
while(abs(cor(X1, Y1)) > 0.002){
  X1 <- rnorm(n = 24, mean = 0, sd = 1);
  Y1 <- rnorm(n = 24, mean = 0, sd = 1);
}
plot(x = X1, y = Y1, pch = 20, xlab = "X value", cex = 1.5, xlim = c(-3, 3),
     ylim = c(-3, 3), xaxt = "n", yaxt = "n", cex.lab = 1.25, ylab = "Y value",
     line = 1);
```


## Getting a more intuitive sense of correlation


Formalised with the **correlation coefficient** (*r*)

- Provides a statistical measure of strength and direction of correlation
- Only describes association between variables (**not** cause and effect)

The value *r* ranges between -1 and 1

- Negative numbers indicate a negative correlation
- Positive numbers indicate a positive correlation
- Avalue of zero indicates no correlation

We can get a more intuitive understanding of the correlation coefficient with **[[this application](https://bradduthie.shinyapps.io/corr_click/)]**.


